Likelihoods for a general class of ARGs under the SMC

### 2025-04-23 review-responses.tex

\begin{point}{\revref}
I would appreciate if you here make it clear which information is missing in these different tools. Sorry for nitpicking.
\end{point}

\reply{
This is quite a subtle point which we feel would distract from the main
message here by expanding. We have attempted to clarify by stating
that a sequence of local trees does not have enough information to
reconstruct a full ARG (required for likelihood calculation under KYF)
and referred to Wong et al for the details, where there is a full
section dedicated to the correspondence between an  ARG and the
sequence of local trees.
}

\begin{point}{\revref} % 98:
 a new formulation of SMC model and corresponding ARG likelihood
\end{point}

\reply{
We are hesitant to call this a ``new formulation of the SMC''
because we are very much building on the foundations laid by
McVean and Cardin. They developed the SMC initially in a
backwards-time context, but the field has subsequently
focused on the left-to-right perspective. Thus, we wish
to keep this as-is.
}

### 2025-04-23 paper.tex
Some responses

% % % % % % % % % % % % % % % % % %
\subsection*{ARGs} \label{par:recording}
Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labelled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals, or \emph{samples}.
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarised as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome, \revpoint{2}{14}
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome. \revpoint{2}{15}

@@

An ARG can therefore be thought of either as a collection of inheritance
relationships between ancestral haplotypes, \revpoint{2}{16}
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $A$
inherited the left half of its genome from ancestor $g$
and the right half from ancestor $f$
is seen in the fact that the first two trees have $A$ below $g$,
and the remaining have $A$ below $f$.
The blue dotted line from $A$ in the second tree to $f$ in the third tree
draws our attention to the fact that
the path traced by $A$'s ancestry switches at that point.
In fact, $A$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labelled by black numbers) are either samples
or \emph{common ancestry} nodes -- \revpoint{2}{20}
ancestors that are the most recent common ancestor for two
or more samples
at some place along the genome.

>>>>

% % % % % % % % % % % % % % % % % %
\subsection*{ARGs} \label{par:recording}
Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labelled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals, or \emph{samples}.
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarised as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific genome, \revpoint{2}{14}
and that the ARG specifies the genealogical tree describing how the
sampled genomes are related
at each base pair. \revpoint{2}{15}

@@

An ARG can therefore be thought of either as a collection of inheritance
relationships between
sampled and ancestral haplotypes (haploid genomes), \revpoint{2}{16}
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $A$
inherited the left half of its genome from ancestor $g$
and the right half from ancestor $f$
is seen in the fact that the first two trees have $A$ below $g$,
and the remaining have $A$ below $f$.
The blue dotted line from $A$ in the second tree to $f$ in the third tree
draws our attention to the fact that
the path traced by $A$'s ancestry switches at that point.
In fact, $A$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labelled by black numbers) are either samples
or \emph{common ancestry} nodes -- \revpoint{2}{20}
ancestors that are the most recent common ancestor for two
or more samples
at some place along the genome.
